---
title: "Mututal Exclusivity"
author: "Sumana Srivatsa"
date: "9/13/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

```

## Data preprocessing 

The `runTimex` function processes the mutation and copy number data to build the alteration matrix for COAD and READ for MSS and MSI samples, respectively. Note that we run TiMEX in this function but it is a dummy run for convenience (helps combine the COAD and READ matrices in the next step).

```{r preprocess}
library(TiMEx)
library(dplyr)
library(biclust)

CN_thresh <- 1
types     <- c("COAD", "READ")  #"COAD" "READ"
p_val_thr <- 0.1      

runTimex <- function(type, p_val_thr = 0.05, ms_status){
    # Load mutation data
    load(paste("./Data/mutations_TCGA-",type, ".RData", sep =""))
    # Load mss status
    load(paste("./Data/TCGA-",type,"_msi_mss.rda", sep = ""))
    # Read CNV data
    all_CN_mat <- read.delim2(paste("./Data/TCGA-",type,"_Gistic2_CopyNumber.by_genes", sep =""),stringsAsFactors = FALSE)
  
    # Binarize the copy number matrix with values -2 or + 2 set to 1 and 0 otherwise 
    all_CN_mat <- data.frame(all_CN_mat[,-1], row.names=all_CN_mat[,1])
    all_CN_mat <- binarize(abs(all_CN_mat), threshold = CN_thresh)
  
    # Set of drivers and TGF regulators
    drivers <- c("ARID1A", "ARID2", 
                 "SMARCA4", "SMARCB1", 
                 "CNIH4", "KEAP1", 
                 "FAM122A", "NBAS", 
                 "CDKN2A", "CDKN2B")
    tgf_reg <- c("TGFBR1", "TGFBR2", 
                 "SMAD4", "SMAD3")
    
    # Reformatting the colnames of mutation matrix to match with copy number data
    patients   <- sapply(colnames(mut_data$missense),
                         function(x){paste(strsplit(x,"[-]")[[1]][1:4],collapse = ".")})
    patients   <- sapply(patients, function(x){substr(x, 1, nchar(x)-1)})
    colnames(mut_data$missense)   <- patients
    colnames(mut_data$truncating) <- patients
    # Common samples between mutation and copy number data
    patients   <- intersect(patients, colnames(all_CN_mat))
  
    # categorizing patients based on MSS or MSI status
    if(ms_status == "MSS"){
      ms_samples <- msires$bcr_patient_barcode[msires$mononucleotide_and_dinucleotide_marker_panel_analysis_status == ms_status] 
    
      # The sample barcodes are different between msires and patients, 
      # need to convert them temporarily
      temp_patients <- patients
      temp_patients <- sapply(temp_patients,
                              function(x){paste(strsplit(x,"[.]")[[1]][1:3],collapse = "-")})
      patients      <- patients[which(temp_patients %in% ms_samples)]
    }else if(ms_status == "MSI"){
      ms_samples <- msires$bcr_patient_barcode[msires$mononucleotide_and_dinucleotide_marker_panel_analysis_status %in% c("MSI-H", "MSI-L", "MSI")] 
      # The sample barcodes are different between msires and patients, 
      # need to convert them temporarily
      temp_patients <- patients
      temp_patients <- sapply(temp_patients,
                              function(x){paste(strsplit(x,"[.]")[[1]][1:3],collapse = "-")})
      patients      <- patients[which(temp_patients %in% ms_samples)]
    }else{
      ms_samples <- NULL
    }
    cat("The number of samples with", ms_status, length(patients))
    # Missense, truncated mutations and copy number for driver gene
    # This step of initializing matrix of 0s is important as mutation informations vary for genes
    miss_mat <- trun_mat <- CNA_mat <- matrix(0, nrow = length(c(drivers,tgf_reg)), ncol = length(patients))
    all_genes <- c(drivers,tgf_reg)
    dimnames(miss_mat)   <- list(all_genes,patients)
    dimnames(trun_mat)   <- list(all_genes,patients)
    dimnames(CNA_mat)    <- list(all_genes,patients)
    
    # For the missense mutations
    miss_mat[intersect(all_genes, rownames(mut_data$missense)), patients]    <- mut_data$missense[intersect(all_genes, rownames(mut_data$missense)),patients]
    # For the truncated mutations
    trun_mat[intersect(all_genes, rownames(mut_data$truncating)), patients]  <- mut_data$truncating[intersect(all_genes, rownames(mut_data$truncating)),patients]
    # For copy number data
    CNA_mat[intersect(all_genes, rownames(all_CN_mat)), patients]            <- as.matrix(all_CN_mat[intersect(all_genes, rownames(all_CN_mat)),patients])
  
    # Binarizing matrix for TiMEx mut_mat without CNA and comb_alterations is with CNA
    mut_mat          <- binarize(miss_mat + trun_mat, 0)
    comb_alterations <- t(binarize(miss_mat + trun_mat + CNA_mat, 0))
    rownames(mut_mat)<- paste(rownames(mut_mat),"mut",sep="_")
    rownames(CNA_mat)<- paste(rownames(CNA_mat),"CNA",sep="_")
    all_alterations  <- cbind(t(mut_mat),t(CNA_mat))  
    
    # Dummy runs of TiMEx with CNA and mutations separate and combined
    result_all      <- TiMEx(all_alterations, pairPvalue = p_val_thr)
    result_comb     <- TiMEx(comb_alterations, pairPvalue = p_val_thr)
  
  return(list(all = result_all, comb = result_comb))
}


```

## TiMEx Run

Next for each category of MS (MSI, MSS or none), we run the `runTimex` function for COAD and READ samples. We can either use combined mutation matrix where a gene in a sample has a value of 1 if it is subject to non-synonymous mutations and/or CNA, or we can use mutation matrix where a gene in a given sample has two disctinct entries, i.e, CNA and non-synonymous mutations are treated separately. Nex run TiMEx for each of these mutation matrices at FDR = 10%. 

```{r Timex}
path2res <- "~/Downloads/"

ms_types <- c("MSS","MSI","None")

for(ms_status in ms_types){
  # Execute the  runTimex function before this step
  result              <- sapply(types, 
                                runTimex, 
                                p_val_thr = p_val_thr, 
                                ms_status = ms_status, 
                                simplify = F)
  
  # Combining matrices for COAD and READ before running TiMEx
  # All: gene_mut and gene_CNA are treated differently
  # Comb: mutations and CNA of a gene are combined as single alterations
  comb_alterations    <- do.call(rbind, sapply(result, function(x){x$comb$matrix}))
  all_alterations     <- do.call(rbind, sapply(result, function(x){x$all$matrix}))
  fin_result_comb     <- TiMEx(comb_alterations, pairPvalue = p_val_thr)
  fin_result_all      <- TiMEx(all_alterations, pairPvalue = p_val_thr)
  
  cat("Total number of samples with", ms_status, dim(all_alterations)[1],"\n")
  cat("Total number of samples with", ms_status, dim(comb_alterations)[1],"\n")
  
  # Writing results of combined alterations
  temp_comb <- cbind.data.frame(fin_result_comb$genesSignif[[2]]$fdr,
                                fin_result_comb$pvals[[2]]$fdr)
  colnames(temp_comb) <- c("GeneA", "GeneB","adj_pval")
  write.table(temp_comb, paste(path2res, "COADREAD_CNA_comb", "_", ms_status, "_", p_val_thr, "_TiMEx.txt", sep = ""),quote = F, sep = "\t", row.names = F, col.names = T)
  
  # Writing results of 'all' alterations
  temp_all <- cbind.data.frame(fin_result_all$genesSignif[[2]]$fdr,
                               fin_result_all$pvals[[2]]$fdr)
  colnames(temp_all) <- c("GeneA", "GeneB","adj_pval")
  write.table(temp_all, paste(path2res, "COADREAD_CNA_all", "_", ms_status, "_", p_val_thr, "_TiMEx.txt", sep = ""),quote = F, sep = "\t", row.names = F, col.names = T)
}

```
